Gametic Selection, Meiotic Drive, Sex Ratio Bias, and Transitions Between Sex Determination Systems

### 2017-02-06 LaTeX/Turnover.tex
sunday work on general order for matt version

\subsection*{Turnover between sex-determination systems} 
The evolution of a new sex determination system requires that a rare mutant allele, $m$, at the novel sex-determining locus  increases in frequency when rare. 
The spread of a rare mutant $m$ at the \textbf{M} locus is determined by the leading eigenvalue, $\lambda$, of the system described by the next generation frequency of eggs and sperm carrying the mutation, (\ref{eq:recursions}c), (\ref{eq:recursions}d), (\ref{eq:recursions}g), (\ref{eq:recursions}h), which is an eight equation system. %, evaluated at the above equilibrium.
%Thus, the system describing the change in frequency of the $m$ allele consists of eight recursion equations. 
Dominant neo-Y (when $k=0$) or neo-W alleles (when $k=1$) are only found in male diploids (neo-Y) or female diploids (neo-W) such that their growth rate ultimately depends only on the change in frequency of $m$-bearing gametes produced by males (for a neo-Y) or by females (for a neo-W). 
Furthermore, if the $m$ allele is fully dominant over the ancestral sex-determining system, phenotypes are not affected by the genotype at the ancestral sex-determining region (\textbf{X} locus). 
Therefore, the invasion of rare dominant neo-Y or neo-W alleles depends on only two equations and is therefore determined by the largest eigenvalue that solves the quadratic characteristic polynomial
$\lambda^2+ b \lambda + c =0$.
%\begin{equation}
%\lambda^2+ b \lambda + c =0
%\label{eq:charpoly_neoY}
%\end{equation}
%
%\noindent
%where $b$ is the average of the growth rates of the two haplotypes that carry the $m$ allele ($mA$ and $ma$), $b=(\lambda_{mA}+\lambda_{ma})/2$, and $c$ also involves the fitness of $m$ alleles when they recombine onto the other \textbf{A} background in a heterozygote, $c=\lambda_{mA}\lambda_{ma}+\rho_{mA}\rho_{ma}$ (see table \ref{tab:haplotype_growth}).
In this case $b= - (\lambda_{mA} + \lambda_{ma})$ and $c = \lambda_{mA} \lambda_{ma} -\rho_{mA} \rho_{ma}$, where $2\lambda_{mi}-1$ is the growth rate of mutant haplotypes on background $i\in\{A,a\}$, accounting for loss due to recombination, and $\rho_{mi}-1$ is the rate of addition of mutant haplotypes onto background $i\in\{A,a\}$ due to recombination (see table \ref{tab:haplotype_growth}).
\textcolor{red}{check these interpretations as there are now two 2s in the lambdas of table 2. perhaps the 2 in the denominator is needed because the frequency of males sums to one as does the frequency of females? the two in the numerator is from heterozygotes (like Hardy-Weinberg)?}

@@

Table \ref{tab:haplotype_growth} illustrates a number of key points about the invasion of neo-Y and neo-W mutations. 
For a neo-Y, invasion depends on the relative lifetime fitness of $A$-bearing and $a$-bearing male gametes (i.e., sperm only).
The lifetime fitness of male gametes partly depends on the allele carried by the female gamete that they mate with (e.g., $A$ with probability $p_{X}^{\female} w_{A}^{\female}/\bar{w}_{H}^{\female}$).
Similarly, invasion of a neo-W depends on the relative lifetime fitness of $A$-bearing and $a$-bearing female gametes (i.e., eggs only). 
However, in the case of a neo-W, the allele carried by the male gamete that they mate with can come from either an X-bearing or a Y-bearing sperm (e.g., $A$ with probability $\bar{p}^{\male} w_{A}^{\male}/\bar{w}_{H}^{\male}$). 
In either case, the zygote will then develop as a female due to the presence of a neo-W. 
By contrast, females that do not carry the neo-W only result from matings with X-bearing sperm (e.g., matings with $A$-bearing sperm occur with probability $\bar{p}_{X}^{\male} w_{A}^{\male}/\bar{w}_{H}^{\male}$). 
If the \textbf{A} locus is initially linked to the ancestral sex-determining locus, \textbf{X}, (i.e., $r<1/2$) the frequency of the $A$ allele among X- and Y-bearing sperm can differ (equation \ref{eq:freq_diffs}). 
Thus, eggs with and without a neo-W differ in the frequency of $A$ alleles they obtain from mating with male gametes.

>>>>

\subsection*{Turnover between sex-determination systems} 
The evolution of a new sex determination system requires that a rare mutant allele, $m$, at the novel sex-determining locus  increases in frequency when rare. 
The spread of a rare mutant $m$ at the \textbf{M} locus is determined by the leading eigenvalue, $\lambda$, of the system described by the next generation frequency of eggs and sperm carrying the mutation, (\ref{eq:recursions}c), (\ref{eq:recursions}d), (\ref{eq:recursions}g), (\ref{eq:recursions}h), which is an eight equation system. %, evaluated at the above equilibrium.
%Thus, the system describing the change in frequency of the $m$ allele consists of eight recursion equations. 
Dominant neo-Y (when $k=0$) or neo-W alleles (when $k=1$) are only found in male diploids (neo-Y) or female diploids (neo-W) such that their growth rate ultimately depends only on the change in frequency of $m$-bearing gametes produced by males (for a neo-Y) or by females (for a neo-W). 
Furthermore, if the $m$ allele is fully dominant over the ancestral sex-determining system, phenotypes are not affected by the genotype at the ancestral sex-determining region (\textbf{X} locus). 
Therefore, the invasion of rare dominant neo-Y or neo-W alleles depends on only two equations and is therefore determined by the largest eigenvalue that solves the quadratic characteristic polynomial
$\lambda^2+ b \lambda + c =0$.
%\begin{equation}
%\lambda^2+ b \lambda + c =0
%\label{eq:charpoly_neoY}
%\end{equation}
%
%\noindent
%where $b$ is the average of the growth rates of the two haplotypes that carry the $m$ allele ($mA$ and $ma$), $b=(\lambda_{mA}+\lambda_{ma})/2$, and $c$ also involves the fitness of $m$ alleles when they recombine onto the other \textbf{A} background in a heterozygote, $c=\lambda_{mA}\lambda_{ma}+\rho_{mA}\rho_{ma}$ (see table \ref{tab:haplotype_growth}).
In this case $b= - (\lambda_{mA} + \lambda_{ma})$ and $c = \lambda_{mA} \lambda_{ma} -\rho_{mA} \rho_{ma}$, where $\lambda_{mi}-1$ is the growth rate of mutant haplotypes on background $i\in\{A,a\}$, accounting for loss due to recombination, and $\rho_{mi}-1$ is the rate of addition of mutant haplotypes onto background $i\in\{A,a\}$ due to recombination (see table \ref{tab:haplotype_growth}).
\textcolor{red}{check these interpretations as there are now two 2s in the lambdas of table 2. perhaps the 2 in the denominator is needed because the frequency of males sums to one as does the frequency of females? the two in the numerator is from heterozygotes (like Hardy-Weinberg)? should we absorb it into the drive or recombination term?}

@@

Table \ref{tab:haplotype_growth} illustrates a number of key points about the invasion of neo-Y and neo-W mutations. 
For a neo-Y, invasion depends on the relative lifetime fitness of $A$-bearing and $a$-bearing male gametes (i.e., sperm only).
The lifetime fitness of male gametes partly depends on the allele carried by the female gamete that they mate with (e.g., $A$ with probability $p_{X}^{\female} w_{A}^{\female}/\bar{w}_{H}^{\female}$).
Similarly, invasion of a neo-W depends on the relative lifetime fitness of $A$-bearing and $a$-bearing female gametes (i.e., eggs only). 
However, in the case of a neo-W, the allele carried by the male gamete that they mate with can come from either an X-bearing or a Y-bearing sperm (e.g., $A$ with probability $\bar{p}^{\male} w_{A}^{\male}/\bar{w}_{H}^{\male}$). 
In either case, the zygote will then develop as a female due to the presence of a neo-W. 
By contrast, females that do not carry the neo-W only result from matings with X-bearing sperm (e.g., matings with $A$-bearing sperm occur with probability $\bar{p}_{X}^{\male} w_{A}^{\male}/\bar{w}_{H}^{\male}$). 
If the \textbf{A} locus is initially linked to the ancestral sex-determining locus, \textbf{X}, the frequency of the $A$ allele among X- and Y-bearing sperm can differ (equation \ref{eq:freq_diffs}). 
Thus, eggs with and without a neo-W differ in the frequency of $A$ alleles they obtain from mating with male gametes.
